---
title: "Crime Rates"
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```



### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv

```{r}
crimes <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv") |> clean_names()
```


### Task 3: Use janitor to clean up the column names



### Task 4: Calculate the total crime rate for each year (picking our unit!)

```{r}
crimes |> mutate(total_rate = grand_total/population * 10000)
```



### Task 5: Calculate the violent and property crime rates for each year (same unit)

```{r}
crimes_with_rates <- crimes |> mutate(violent_rate = violent_crime_total/population * 10000, property_rate = property_crime_total/population * 10000)
```



### Task 6: Describe our results

Write a few sentences about what you found in the data. What years had the highest crime rates? What years had the lowest crime rates? Does the pattern match your expectations?

Violent crime rate had a big spike in the early 90s and it slightly coincides with an increase in property crime rates during the same time period. However, the violent crime rates reach beyond what they were in 1975 at the beginning of the data. The property crime rates do not exceed their 1975 number. Overall, both rates do decrease over the course of the last 45 years with violent crimes having a steady rate over the latest 10 years. The latest 2-3 years saw  a big drop off for the property crime rate. This does match my expectations because in recent years, alongside technological advancements, policing has become more strict and organized, allowing significantly less crimes to occur. The rise in a higher violent crime rate in the early 1990s may be do to the political climate at the time or perhaps the rise of the internet around that time, giving people more access to news and the goingson around the world.



When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

