---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Austin DeSisto"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

There is no 100% values on here which is a good start. The data is much more compact here and shows more of a uniform drop than the dratic 100 -> 54 we saw in the PA data. I am still concerned about the relationship between these two variables because I see a wid variety of percentages on the math scores side. 


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

There is a much more clear direction with this scatterplot. It is a slight downward trend and looks to be fairly strong in its angle although there are some datapoints that fall far outside the general direction of the line. One possible explanation for the difference in scatterplots is the reporting of the percentage of FARMS eligible students or the requrements that need to be met might be different for each county. 

### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

p-value = < 0.00000000000000022
adjusted r-squared = 67.34%
coefficient =  -0.59511

All of these values show a statistically significant and slightly strong association between the two variables. While not super close to their maximums, the adjusted r-squared and the coefficient are above halfway there and show that the FARMS pct and math scores have some sort of decent effect on each other. 

### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model_filtered, newdata = schools),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

note on above: i used ChatGPT to discover the error showing 49 rows instead of 127. it suggested that i had the newdata = schools line/

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

Kemp Mill Elementary - largest negative residual.
- The school is underperofmring by a lot and is way under its predicted value. There is a very high hispanic population and the income per capita is around $49k. I would ask, "Why have you been seeing a low number of students meeting or exceeding expectations in math tests?

Cold Spring Elementary largest positive residual.
- The school is overperforming by a fair amount. It has a high white population and a fair asian population totalling almost a quarter or the town. the per capita income is around $114k. I would ask, "How have you been able to see such success in standardized math test scores?"


### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

The data that we have from the Maryland county supports a much stronger relationship between the two variables which back up the board member's claim. It appears that some of the data from PA might be misreported or inaccurate. It does not support the cliam very well at all. I belive this could also be because of the fewer number of schools in the PA county. I would really want to visit the Cold Spring Elementary school because it is overperforming by so much that there must be some other reasons. There is a point of the story here about why one state and county support the claim much stronger than the other one and I believe a further investigation could be of benefit.  
