---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

Many of these incidents have a few different breakups of their categories. For example B&E is split up into the type of building that it occurred in. Form a reporting standpoint, I think that the "Auto Stolen" categories are most interesting because there appears to be a slight trend and you can potentially find a correlations between stolen and recovered."Robbery Vehicle" and "Theft from Auto" seem a bit confusing as they seem to indicate a similar crime but the dataset indicates no further description.

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |>
  mutate(total_incidents = Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec, average_incidents = total_incidents / 12)
```


### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The categories with the highest average number of incidents per month is accident, auto stolen and theft from auto. The categories with the lowest are B&E other and homicides. It seems that there is constantly a large problem with crimes related to cars and personal vehicles. It seems that overall throughout the year, the crimes maintain a relatively consistent level with few outliers.


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> mutate(june_change = (Jun - average_incidents)/average_incidents * 100,
  july_change = (Jul - average_incidents)/average_incidents * 100,
  august_change = (Aug - average_incidents)/average_incidents * 100
)
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

In the auto stolen and recovered category, the percent change in the average sees a very large increase in july preceded by an increase in june but then a slight decrease in the month of august. Another category, homicide, saw a jump of 37% in june and then smaller increases in the rest of the summer months. Given this data, it does appear that overall a lot of the crimes tend to see an increase in the summer months while some do tend to taper off toward the end of the summer.

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

A 25% spike in commercial breaking and entering was recorded in August in Prince Georges County, according to new data. 

The increase is compared to the average over the course of the year, which sits at 44.6. In the first month of summer, there were 48 commercial breaking and entering crimes reported. The last month of the summer, August, saw 51 of the crime category reported. While each month had an above or near average reporting, the months had more break-ins during the summer than any other time of year, excluding December. The massive spike in December could be due to the greater need for warmth or shelter in the winter months. In contrast, the increase in the summer could be the need for cooler temperatures in the blistering heat.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
