---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

Overall, there is a wide variety of salaries ranging from the low 60k range to the near or above 200k range. Most of the people with "Special Assistant" titles can vary in their salaries from the 120k to 135k range. It might be hard to determine exactly why there is such variability. Finally, there are a few staffers with a reported salary of 0. It is unclear why this is and there is seemingly no pattern besides vage titles such as special assistant and senior advisor.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}

wh_salaries |> summarize(
  
  mean(SALARY),
  median(SALARY)
  
  
)

```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}

wh_salaries |> summarize(
  
  mean_salary = mean(SALARY),
  median_salary = median(SALARY),
  min_salary = min(SALARY),
  max_salary = max(SALARY)
  
  
)

```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

As I stated earlier, the minimum salary I observed is 0. This is most certainly an outlier because it would not make sense for someone to be paid nothing to work at the White House. The mean and median are actually not very far off, with the median being lower than the mean indicating that the data is skewed right. This shows that there are more salaries in the higher range than in the lower range, even despite the mean being pulled down by the 0s.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

My best story idea would be find the people that it in the lowst parts of the salaries and discover why they are so low. I decent next step would be to try and contact the White House and those people to see if we could get insight into what their job might look like. 
