---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(
  
  violent_rate_2019 = violent2019/pop2019 * 10000,
  property_rate_2019 = property2019/pop2019 * 10000,
  violent_rate_2020 = violent2020/pop2020 * 10000,
  property_rate_2020 = property2020/pop2020 * 10000
  
  
  
  )
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

An overall pattern I observed is that places in Maryland with higher population densities such as Baltimore City and Prince Georges County have overall higher rates of both crimes across all years. Focusing specifically on the two different types of crime, Baltimore City has a higher property crime rate than violent crime rate, but both are in the tripple digits. However, there is a larger discrepancy between the two in a place like Harford County. The story might be discovering why some counties have higher violent crime rates vs others that are very low. 

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities |> mutate(
  
 change_violent = (violent2020 - violent2019) / violent2019 * 100,
 change_property = (property2020 - property2019) / property2019 * 100
  
  
)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

What is most interesting about the percent change is that cities with smaller populations seem to have much larger percent change numbers than those with larger populations. This makes sense because even a small change in crime, when not many crimes have been committed, will result in a large percent change. Overall, violent crime saw an increase year over year while property crime saw a general decrease. Part of his may be due to the onset of covid, where less people were outside and keeping interactions to a minimum. The story might be trying to really determine the validity of the data given the pandemic that encompassed 2020.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Maryland cities saw a general decrease in property crime from 2019 to 2020, as the coronavirus swept through the state.

Most cities in Maryland saw at least a 50% decrease in property crime year over year from 2019 to 2020. One of the most obvious reasons may be the lockdown that much of the world was under during the greater part of 2020. The lack of social inteaction and being in public meant that there was simply less opportunity to commit those property crimes. Some cities even saw an over 100% decrease in property crimes year over year.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: Percentiles


```{r}
md_cities_percentiles <- md_cities_rates |> mutate(percentile = ntile(violent_rate_2020, 100))
  
  
```

There are 11 cities above the 90th percentile.

```{r}
md_cities_percentiles |> summarize(
  
  IQR = IQR(violent_rate_2020)
  
  
)
```



Yes, I think it changes the story idea a little bit because it helps compare where all of the cities are at. I still think it would be advantageous to look at another set of two years pre-covid to provide a nice comparison.
