---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```



### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
mandms <- read_csv('https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv')
```



### Task 3: Sample 5 rows of data from the dataset

```{r}
sample_5 <-  mandms|> sample_n(5)
```



### Task 4: Calculate the mean age and standard deviation of the sample, using either the blue or white number of M&Ms

```{r}
mean_sample_red = mean(sample_5$red)
sd_sample_red = sd(sample_5$red)

mean_sample_red
sd_sample_red
```



### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4

```{r}
mean_pop_red = mean(mandms$red)
sd_pop_red = sd(mandms$red)

mean_pop_red
sd_pop_red
```



### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

They are similar, but the sample results are giving considerably higher numbers than the actual mean and sd from the population. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

The very small sample size revealed a large variation in the red totals which is not necessarily an accurate representation of the entire population. Such a small set of the population is unlikely to yield super accurate numbers, but given that is is random, should be able to get pretty close, which it does. 


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
