---
title: "final_project"
output: html_document
DATA THROUGH APRIL 22, 2026
---

The two issues I am running into after the tests are the error shown below with mismatching rows and the issue of it not counting all of the rows in the group in the second code block down. I'm not sure exactly what teh fix might be here but i suspect there's a way to make sure it counts each row. As for the below error, I'm sure there might also be a way to have it show 2018 and 2019 as 0, but again, not sure how to do that and I didn't get a chance to scour the internet. 



```{r}
by_year <- cbind(by_year_up, by_year_down)

by_year |>
  ggplot(aes(x = `mean_new_sats`, y = `mean_only_down`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs( x="Total Starlink Satellites Launched", y="Satellites Not in Orbit", title = "Total launched Starlinks vs. Starlinks Not in Orbit", caption = "Jonathan McDowell, https://planet4589.org")
```

```{r}
library(dplyr)

by_year_up <- not_zeros %>% group_by(across(Year)) %>%
  summarise(
    mean_new_sats = mean(new_sats, na.rm = TRUE),
    total_count_new = n(),
    total_launch = sum(new_sats)
  
  )

by_year_down <-  only_down %>% group_by(Year) %>%
  summarize(
    mean_only_down = mean(new_down, na.rm = TRUE),
    total_count_down = n(),
    total_down = sum(new_down)
  )

by_year_up
by_year_down


```

Not sure why im not getting all of the rows accounted for. Used (https://builtin.com/data-science/grouping-r) for help on this.

LINEAR REEGRESSION BELOW

```{r}
model <- lm(`Down` ~ `Tot`, data = starlink)
summary(model)
```
I'm truly not sure how some of these numbers are meaning in this context. I can see that the adjusted R-squared is 0.9167 which shows a very strong correlation between the total satellites launched and the satellites not in orbit. This data is pulling from the totals, which is ok for this sort of plot. The p-value is super super low which shows that it is statistically significant, but I can't determine this coeficient and I'm not sure why its so low. I believe this shows the there is a large effect on the number of satellites down as the overall population increases. There is simply a high chance that any given satellite has an issue. What's interesting is the point that it sees a big spike in downed satellites just aver 7500 total. A few explanations for this could be that they started launching a new satellite variant which experianced more issues and then the started to figure out what the issue was, as indicated bt the flattening of the line again.

```{r}
starlink |>
  ggplot(aes(x = `Tot`, y = `Down`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs( x="Total Starlink Satellites Launched", y="Satellites Not in Orbit", title = "Total launched Starlinks vs. Starlinks Not in Orbit", caption = "Jonathan McDowell, https://planet4589.org")
```


```{r}
only_down <- starlink |> filter(new_down != 0)
```

```{r}
starlink <- starlink |> 
  arrange(Date) |> 
  mutate(
    new_sats   = `Tot` - lag(`Tot`),
    new_down   = `Down` - lag(`Down`),  # adjust column name to whatever read_table assigns
    new_dead   = `M` - lag(`M`)
  )

not_zeros <- starlink |> filter(new_sats != 0)
```

```{r}


starlink <- read_table(
  "starlink_log.dat",
  comment = "#",
  col_names = c("Bin", "Date", "Tot", "FTO", "FD", "RD", "MD", 
                "Down", "Up", "F", "M", "G", "Work", "R", "L", 
                "U", "T", "S", "D", "A", "O",
                "Year", "Month", "Day", "Time")
)

starlink

```


```{r}
library(tidyverse)
library(lubridate)

readLines("starlink_log.dat", n = 20)
```


```{r}
library(tidyverse)

url <- "https://planet4589.org/space/con/star/log.dat"

starlink_data <- read_fwf(
  file = url,
  skip = 2,
  col_positions = fwf_widths(
    c(4, 8, 5, 4, 4, 4, 4, 6, 4, 3, 3, 3, 6, 3, 3, 3, 3, 3, 3, 3, 3),
    col_names = c(
      "row_num","Bin","Date","Tot","FTO","FD","RD","MD",
      "Down","Up","F","M","G","Work","R","L","U","T","S","D","A"
    )
  )
)

starlink_data
```

```{r}
not_zeros <- data |> filter(new_sats != 0)
```


Note: use ChatGPT (https://chatgpt.com/share/69ee9b6e-e3bc-83ea-a5d1-bc28a9b4c644) to help me understand the code shift issue and change a bit of what you had done to get the table looking right! (but it hasn't worked very well) 


List of potential analyses:
1. scatterplot to see correlation of total number of satellites launches vs downed satellites 
2. mean number of satellites launched per month and see if there are any outlier months. this will be interesting because we might have to figure out how to negate the increase in cadence
3. frequency of new satellites added over time
4.how often there is a downed satellite and the trend of downed satellites over time
5. linear regression of new sats vs down sats

I'm not sure if there's more specific tests to run but those are generally what I think could help answer my hypothesis.

Note: somehow I think the below code shifted the table to the left by one column. So the original 'Date' reads the 'Tot' column. I tried to investigate but I do believe I am clueless.

```{r}
library(tidyverse)
library(lubridate)

raw <- read_lines("https://planet4589.org/space/con/star/log.dat")

# Find the actual header row
header_i <- which(str_detect(raw, "^\\s*#\\s+Bin\\s+Date\\s+Tot"))[1]

header_line <- raw[header_i] |>
  str_remove("^\\s*#\\s*")

data_lines <- raw[(header_i + 1):length(raw)]

starlink_data <- read_table(
  I(c(header_line, data_lines)),
) |>
  mutate(Date = ymd(Date))

starlink_data
```
s


```{r}
starlink_data <- starlink_data |> 
  arrange(Date) |> 
  mutate(
    new_sats   = `Tot` - lag(`Tot`),
    new_down   = `Down` - lag(`Down`),  # adjust column name to whatever read_table assigns
    new_dead   = `M` - lag(`M`)
  )
```




I was able to load the table using your first suggestion! 

HYPOHTESIS: My hypothesis is that more satellites in the constellation would lead to more frequent "downing" of satellites due to the fact that more satellites equals more opportunities for any single satellites to fail. There is also the potential that some of the downed satellites may relate to significant solar events or launch failures/anomalies which could be newsworthy as well.


Data explanations as given by the source website (https://planet4589.org/space/con/star/log.html): 

Sats: Total number of satellites launched.
Failed to orbit (f): Launch vehicle failed to reach orbit.
Down (F) Early deorbit: satellites which have reentered; determined to be malfunctioning before reaching operational orbit, and abandoned or actively lowered prior to reentry.
Down (R) Disposal, later semicontrolled reentry: satellites which reached the operational orbit but whose orbits were later lowered to cause early reentry. These satellites retained propulsive capability but either failed in some other respect or were deliberately retired.
Down (M) Reentry after fail; Later uncontrolled reentry: Reentered after an extended period of uncontrolled orbital decay; presumed failed.
Total down: Total of satellites which were launched but are not currently in orbit.
Total in orbit: Total of satellites which reached orbit and have not yet reentered, whether working or not.
Screened (F): Satellites not working correctly and left in low orbit prior to completing orbit raising; expected to reenter soon. Includes satellites that failed at launch.
Dead (M): Failed, uncontrolled decay. Satellites stopped stationkeeping after reaching operational orbit; no manuevers occurring, presumed failed.
Graveyard (G): Graveyard orbit, retired. Not used for Starlink.
Total working: Total of satellites still in orbit minus those thought failed (screened or uncontrolled decay).
Retiring (R): Disposal. Orbit lowering to semicontrolled reentry underway
Lowered (L): Out of con(stellation). Satellites which were operational but have been removed from the operational shell; still actively maneuvering.
Anomalous (U): Satellites lowered to an intermediate orbit height but stationkeeping. I suspect these satellites are not fully operational, but not yet abandoned - perhaps SpaceX hopes to debug their problems.
Reserve (T): Satellites maintained in an orbit close to the operational one, relocating in phase or temporarily out of the constellation. As of mid 2023 I am not bothering to mark T status most of the time for Starlink as a lot of satellites are popping in and out of this status on a daily basis.
Special (S): Satellites making unusual manuevers, suspected to be being used for special tests.
Drift (D): Satellites maintained at an intermediate altitude undergoing targeted orbital plane drift.
Ascent (A): Satellites currently orbit raising towards operational altitude
Op Shell (O): Satellites in one of the operational shells. 

The data goes back to 2018 and goes to the present. Given that I am pulling from the website, we will assume it to be updated to the latest launch. The most important columns will be the 'sats' 'down' and 'dead'. Those are general the most encompassing portions of the data. But there is some other columns that include more parsed out data which may be interesting to discover later on. 

QUESTION: I may ask something similar on the questions assingment, but im notcing that the data page has actual dates but that's not showing up here. Is that something that is wrong with the read_table? Other than that, I'ms working on understanding the letters and titles and will just need to keep track of it as I explore the data more.

