---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Austin DeSisto"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
library(knitr)    
library(lubridate) 
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
health_data <-  read_csv('https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv')
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- health_data |> 
  
  summarize(
    mean_compliance = mean(compliance_score),
    sd_compliance = sd(compliance_score),
    min_compliance = min(compliance_score),
    max_compliance = max(compliance_score),
    )
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
health_data |> 
  ggplot() + 
  geom_histogram(aes(x = compliance_score), binwidth = 3) +
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) +
labs(
    title = "Restuarant Health Compliance Scores",
    subtitle = "Red line = mean compliance scores",
    x = "Score",
    y = "Count",
    caption = "Source: Montgomery County Health Inspection"
  ) +
  theme_minimal()

```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:** The standard deviation of the compliance scores is only 5.8 and the mean sits at 96.3. There is a very clear left skew to the histogram, showing that most of the data is in the 100 or near 100 column. However, a few outliers in the near 60 and 70 range, bring the mean down away from the 100. These are WELL outside of 3 standard deviations. Those are definitely the most news worthy parts and it would be worth looking into why some of these low scoring restaurants got such a low score by reading the reports and maybe visiting them. 

---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}


wmata_daily <-  read_csv('https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv')
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}

wmata_stats <- wmata_daily |>
  
  summarize( 
    mean_bus = mean(bus),
    sd_bus = sd(bus),
    mean_rail = mean(rail),
    sd_rail = sd(rail)
    
    )

nrow(wmata_daily)

```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}

wmata_zscores <- wmata_daily |>
  mutate(
    rail_zscore = ((total - 347987.8) / 100164.3),
    bus_zscore = ((total - 331293.8) / 88486.88)
  )
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
wmata_unusual <- wmata_zscores |>
  filter(abs(rail_zscore) > 2) |>
  arrange(desc(abs(rail_zscore)))

wmata_unusual |>
  select(date, total, rail_zscore) |>
  kable(digits = 2)
```

There are 301 unusual days

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(42)

sample_91 <-  wmata_daily|> sample_n(91)

mean_rail_91 <-  mean(sample_91$rail)
sd_rail_91 <- sd(sample_91$rail)
mean_bus_91 <- mean(sample_91$bus)
sd_bus_91 <- sd(sample_91$bus)

mean_rail_91
sd_rail_91
mean_bus_91
sd_bus_91
```
They are faily close to the actual values I calculted, but are a bit inflated. I think this is because there could be some of the extreme outliers in the data included in the sample.


### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

Given that these days are so far outside of the mean and have such a high z score we could search up events on those days and discover why the ridership might have been so extreme. For example, the Oct. 29 date was a Kamala Harris rally in her run for president, which might have seen extreme ridership difference than the average day.
---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
cartheftsmd <-read_csv('https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv')
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}

md_car_thefts <- cartheftsmd |> 
  
  mutate(
    theft_rate = (thefts_2022 + thefts_2023)/population *10000
    
  )
```

A per-capita rate is more useful because some counties have way higher populations than others and therefore may have more accidents because there are more people driving around in them. This function allows us to compare the counties more easily by leveling them out with a rate per their 10,000 residents.

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
md_car_thefts |> 
  mutate(
    median_statewide = median(theft_rate)
    
)



```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

Generally, counties with a greater population have a greater car theft rate across both years. This is more concentrated in cities where there are more people in a compact area which could lead to more crime and car thefts.
---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)
I would prioritize using a z score because the I would want to know how far off certain pools are. It might not be able to show extreme outliers because the SD is so large, but it would help to get a measurement of exactly how far off they are. Additionally, using a histogram to see this, we can get an idea of where the majority of the data lies. If most of it is below the mean, then there must be many pools that are dramatically underperfomring.

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

I would use z scores to see how far below the mean some of these pools are. Anything outside of -2 is a drastic underperformance but there might be a good cluster that are below -1.5 or nearly could also be considered a big underperformance. Simply ranking the pools is not good enough because some of them might have more people living nearby that frequent them and because it would be hard to compare that to the mean of all the pools which is our measurment for the baseline of the performance of the pools.

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

I would immediately ask about more data on the 60 days that were sampled. There could be many other factors like temperature and stormy weaether that might affect the pools attendance. 60 days is not a large sample at all especially compared to the many months than warm weather can be around, making the sample small compared to the full nudmer of days that data can be gathered for. I would be concerned about the low number of days that were sampled and any specific events from some of the days that might be extreme outliers.

---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
